Exploring probability distributions for bivariate temporal granularities

### 2019-11-18 main.Rmd
distribution section review

<!-- Why distribution plot -->
<!-- Huge data difficult to summaries using one statistic-->
With huge amount of data being available, mean, median or any one summary statistic might not enough to understand a data set. Soon enough following questions become more interesting:

 - Are values clustered around mean/median or mostly around tails? In other words, what is the combined weight of tails relative to the rest of the distribution?

 - Does values rise very quickly between 25th percentile and median but not as quickly between median and 75th percentile? More generally, how the variation is the data set changes across different percentiles/deciles?

  - Is the tail on the left hand side longer than that on the right side? Or are they equally balanced around mean/median?

This is when displaying a probability distribution becomes a potentially useful approach.
The entire distribution can be visualized or some contextual summary statistics can be visualized to emphasize certain properties of the distribution. These properties can throw light on central tendency, skewness, kurtosis, variation of the distribution and can also be useful in detecting extreme behavior or anomalies in the data set.

@@

<!-- Why 2D -->
Graphical output devices are flat. If there are only two variables, a cartesian graph of one variable against the other can show the configuration of points. As soon as the data move to three variables, we need to infer multi-dimensional structure by a two-dimensional medium[Cleveland, elements of graphing data]. Any attempt to encode all these temporal granularities together to develop insights on periodicity might fail or become clumsy. Instead, the big problem can be broken down into smaller pieces by focusing on specific tasks per visual representation in a 2D space.

<!-- Why facet-ing -->
`ggplot2` (Wickham, 2016) is an excellent tool that facilitates the process of mapping different variables to a 2D frame through grammar of graphics. One aspect of ggplot2 is faceting, inspired by Trellis plots (Cleveland, 1993). These are based on conditioning the values taken on by one or more of the variables in a data set. This means creating same plots using data subsets corresponding to each level of the conditioned variable. This is a powerful tool for exploratory data analysis as we can rapidly compare patterns in different parts of the data.

<!-- What are we visualizing -->
We will see the distribution of the time series across bivariate temporal granularities through faceting approach with one temporal granularity plotted along the x-axis, the other one across facets and the dependent time series variable on the y-axis.

Now, different distribution plots might be appropriate depending on which features of the distrbution we are interested to look at, the levels of time granularities being plotted, how granularities interact and number of observations available. We will look at each of these aspects separately and analyze the effect of each on visualization.

@@

<!-- Lit review on statistical distribution plot-->

<!-- Description and lit review of boxplot -->
There are several ways to plot statistical distributions. The displayed probabilities in these plots are either computed using kernel density estimation methods or emperical statistical methods. We will discuss the benefits and challenges of few conventional and recent ways to plot distributions using both of these methods.

### Emperical methods

Most commonly used techniques to display distribution of data include the histogram (Karl Pearson), which shows the prevalence of values grouped into bins and the box-and-whisker plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. The box plot is a compact distributional summary, displaying less detail than a histogram. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg, 1978] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of these variations are helpful to get an idea of the distribution at a glance. However, we do not have an idea if the distribution is multimodal. Moreover, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

<!-- Description and lit review of letter-value plots -->
The letter-value box plot [@Hofmann2017-sg, 2006] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail.It shows only actual data values and no smoothed summaries. However, the letter values are shown till the depths where the letter values are reliable estimates of their corresponding quantiles. These might lead to a lot of letter values being shown and leading to overload of information in one plot. Also, multimodality is difficult to spot.

@@

Quantile plots visually portray the quantiles of the distribution of sample data. Much like the quartiles divide the data set equally into four equal parts, extensions might include dividing the data set even further. These plots are referred to as "quantile" plots with, where number of quantiles 9 for deciles and 99 for percentiles. No distributional assumptions are made about the data since empirical deciles are plotted. It avoids much clutter and just enable us to focus on specific probabilities. These plots do not allow us to see the shape, skewness or nature of the tail of the distribution with so much clarity as violin plots, ridge plots or letter value plots. Also, a large data set is required for the extreme percentiles to be estimated with any accuracy.

@@

A density plot which uses a kernel density estimate to show the probability density function of the variable can show the entire distribution, unlike discretizing the distribution and showing only parts of it.

Violin plots [@Hintze1998-zi, 1998] adds the information available from local density estimates to summary statistics provided by box plots. The shape of the violin represents the density estimate of the variable. The more data points in a specific range, the larger the violin is for that range. Adding two density plots gives a symmetric plot which makes it easier to see the magnitude of the density and compare across categories. Howevr, the density in violin are estimated through kernel density estimation and thus makes assumptions when selecting kernel or bandwidth. As a result, the plot shows smooth summaries based on the assumptions and not only on actual observations.

@@

As a general rule, the quantile plots are really useful for picking patterns, whereas, other more involved methods of plotting are useful for studying anomalies or outlier behavior. It is good to be aware of the benefits and challenges while choosing any distribution plots, depending on the context.

>>>>

<!-- Include why distribution plot, why 2D, why facetting in introduction -->
<!-- Why distribution plot -->
<!-- Huge data difficult to summaries using one statistic-->

<!-- With huge amount of data being available, mean, median or any one summary statistic might not enough to understand a data set. Soon enough following questions become more interesting:   -->

<!--  - Are values clustered around mean/median or mostly around tails? In other words, what is the combined weight of tails relative to the rest of the distribution?  -->

<!--  - Does values rise very quickly between 25th percentile and median but not as quickly between median and 75th percentile? More generally, how the variation is the data set changes across different percentiles/deciles? -->

<!--   - Is the tail on the left hand side longer than that on the right side? Or are they equally balanced around mean/median? -->

<!-- This is when displaying a probability distribution becomes a potentially useful approach. -->
<!-- The entire distribution can be visualized or some contextual summary statistics can be visualized to emphasize certain properties of the distribution. These properties can throw light on central tendency, skewness, kurtosis, variation of the distribution and can also be useful in detecting extreme behavior or anomalies in the data set. -->

@@

<!-- Why 2D -->

<!-- Graphical output devices are flat. If there are only two variables, a cartesian graph of one variable against the other can show the configuration of points. As soon as the data move to three variables, we need to infer multi-dimensional structure by a two-dimensional medium[Cleveland, elements of graphing data]. Any attempt to encode all these temporal granularities together to develop insights on periodicity might fail or become clumsy. Instead, the big problem can be broken down into smaller pieces by focusing on specific tasks per visual representation in a 2D space.   -->

<!-- Why facet-ing -->

<!-- `ggplot2` (Wickham, 2016) is an excellent tool that facilitates the process of mapping different variables to a 2D frame through grammar of graphics. One aspect of ggplot2 is faceting, inspired by Trellis plots (Cleveland, 1993). These are based on conditioning the values taken on by one or more of the variables in a data set. This means creating same plots using data subsets corresponding to each level of the conditioned variable. This is a powerful tool for exploratory data analysis as we can rapidly compare patterns in different parts of the data.  -->

<!-- What are we visualizing -->

<!-- We will see the distribution of the time series across bivariate temporal granularities through faceting approach with one temporal granularity plotted along the x-axis, the other one across facets and the dependent time series variable on the y-axis.  -->

<!-- A well designed visual representation can replace cognitive calculations with simple perpetual inferences and improve speed of comprehension and decision making. However, creating a visualisation requires a number of nuanced judgements. The challenge being given a data set, the number of visual encodings and hence the space of possible visual design is extremely huge.  -->

The grammar of graphics introduced a framework to incorporate underlying features for statistical graphics to construct them by relating data space to the graphic space [@Wilkinson1999-nk]. The layered grammar of graphics proposed by [@Wickham2009-pk], which is a modification over the former, allows us to produce graphics using the same structured thinking that is used to design analysis. Briefly, any graphics are made up of **Layers**, **Facet**,**Scale** and a  **coordinate system**.**Layers** are made up of **data**, **mapping**, **statistical transformation**, **geometric objects** and **position adjustments**.

Creating a visualisation requires a number of nuanced judgements. However, we exploit some features of layers and faceting to come up with a recommendation system while visualizing the distribution of the "dependent" variable across bivariate temporal granularities. The general framework of visualizing the distribution invloves a faceting approach with one temporal granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. Different distribution plots might be appropriate depending on which features of the distrbution we are interested to look at, the levels of time granularities being plotted, how granularities interact and number of observations available. We will look at each of these aspects separately and analyze the effect of each on visualization. While we accomplish that, we need to bear in mind that any customised species of visualization might be constructed by varying the set of mappings between data properties and visual attributes such as position, shape and color.

@@

<!-- Common benefits and challenges of empirical and density based visualization -->

There are several ways to plot statistical distributions. The displayed probabilities in these plots are either computed using kernel density estimation methods or empirical statistical methods. The latter do not allow us to see the shape, skewness or nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus. The probabilities displayed are based on actual data, which is aligned to principles that governed Tukey’s original boxplot. Density plots which uses a kernel density estimate to show the probability density function of the variable can show the entire distribution, unlike discretizing the distribution and showing only parts of it. They can be useful in spotting multimodality, however, they might become obscuring with too many categories and information on the entire distribution can add to cognitive load. Also, the probabilties in density plot are estimated through kernel density estimation and thus makes assumptions when selecting kernel or bandwidth. As a result, the plot shows smooth summaries based on the assumptions and not only on actual observations. The density based visualizations are subject to the same sample size restrictions and challenges that apply to any density estimation. In practice, the densities are estimated reasonably with at least 30 observations. Even with sample sizes of several hundred, however, choosing too large a value for bandwidth can cause the estimate to oversmooth the data.

We will discuss few conventional and recent ways to plot distributions using both of these methods.

### Empirical methods

Most commonly used techniques to display a distribution of data include the histogram (Karl Pearson), which shows the prevalence of values grouped into bins and the box-and-whisker plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. The box plot is a compact distributional summary, displaying less detail than a histogram. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg, 1978] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of these variations are helpful to get an idea of the distribution at a glance. Moreover, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

<!-- Description and lit review of letter-value plots -->
The letter-value box plot [@Hofmann2017-sg, 2006] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail. “outliers” in letter value plots are those observations beyond the most extreme letter value. The letter values are shown till the depths where the letter values are reliable estimates of their corresponding quantiles and hence might lead to a lot of letter values being shown and leading to overload of information in one plot.

@@

Quantile plots visually portray the quantiles of the distribution of sample data. Much like the quartiles divide the data set equally into four equal parts, extensions include dividing the data set even further. These plots are referred to as "quantile" plots. For example, number of quantiles would be 9 for a decile plot and 99 for percentile plot. It avoids much clutter and just enable us to focus on specific probabilities, typically representing typical and extreme behaviors. A large data set is required for the extreme percentiles to be estimated with any accuracy. These plots can display any quantiles instead of quartiles in a traditional boxplot. When reviewing a boxplot, an outlier is defined as a data point that is located outside the fences (“whiskers”) of the boxplot (e.g: outside 1.5 times the interquartile range above the upper quartile and below the lower quartile). However, outliers are open to interpretation and not shown in an quantile plot.

@@

Traditional ways to visualize densities include violin plots [@Hintze1998-zi, 1998]. The shape of the violin represents the density estimate of the variable. The more data points in a specific range, the larger the violin is for that range. Adding two density plots gives a symmetric plot which makes it easier to see the magnitude of the density and compare across categories, enabling easier detection of clusters or bumps within a distribution.

@@

It is good to be conversant with the benefits and challenges while choosing any distribution plot, depending on the context. As a general rule, if we have too many categories, the quantile plots are useful for camparing patterns, whereas, other more involved methods of plotting are useful for studying anomalies, outlier or multimodal behavior.
